Colorimetric Detection of pH Strips

### 2017-12-10 CS189/paper/paper.tex
add period after Fig in-text

We collected our own data for this study. Mutlu was able to capture the range of pH values comprehensively through the use of adjustable chemical buffers. This allowed them to train on chemical solutions with integer pH values. They also were able to determine ground-truth labels for their data points by using a digitized pH meter. We were more limited in our resources; we had a fixed set of non-adjustable buffers and we took the label on the bottle as the ground-truth label for regression. In total we hand-held a smartphone to capture 10 samples under standard household lighting of pH strips submersed in the following 9 buffers: 2.22, 4.40, 4.47, 5.31, 6.28, 7.00, 7.70, 8.70, 10.0 (Fig \ref{raw_photo}). These fixed non-integer pH values may be a source of irreducible error for us when we rounded these values to the nearest integer when performing classification. However, we wanted our model to be more robust to noise and to be applicable in a more reasonable environment, where a researcher would not have to use a custom apparatus remove any external sources of light [1, 2]. Mutlu waited for the strips to dry before taking pictures, but we decided to capture the images while they were still wet since we imagine that pH strips are read while wet in practice.

@@

For our study, as Mutlu did, we extract out the RBG values from the four panels, but we do this as our first step and use these original features for our first experiment. In order to extract the colors, we first crop the original image using the box specified from the horizontal (Fig \ref{pre-process}a) and vertical (Fig \ref{pre-process}b) Sobel edge-detection filters, as illustrated by the red lines in the images. We then apply a repeatedly-opened, distance eroded binary threshold (Fig \ref{pre-process}c) to get the center regions of each color square from the original image (Fig \ref{pre-process}d).

@@

From there, we extract the 12 features by taking the average value RBG values from each of the four color boxes (Fig \ref{four-colors}a). We then generalize Mutlu’s experiments by artificially applying randomly generated lighting conditions to our data. This way, we do not correct for lighting before training a model like Mutlu does but rather, introduce lighting conditions to our homogeneous dataset so that our model can learn to ignore lighting conditions on its own, rather than having to rely on manual and lighting condition-specific pre-processing methods. Specifically, in our second experiment, we gamma correct each sample with a uniformly chosen coefficient between 0.5 and 1.5 (Fig \ref{four-colors}b) because this mimics the nonlinear luminescence introduced in captured images [3]. In our third experiment, we perform hue rotation with a uniformly chosen degree between -90 and 90 degrees (Fig \ref{four-colors}c). In our fourth experiment, we apply the random lighting conditions from both experiments 2 and 3 simultaneously (Fig \ref{four-colors}d).

@@

For select experiments, we used PCA in order to project our data into a lower-dimensional subspace in order to simplify our models and hopefully increase accuracy. We hyperparameter tuned across different numbers of components for the applicable models. We can see that the classes cluster well, even with projection onto only two axes (Fig \ref{pca}).

@@

As in the Mutlu paper, we first used Support Vector Machines (SVM) for classification across the four experiments. Classification accuracies of $98\%$, $96\%$, $88\%$, and $79\%$ were achieved using 7, 8, 9, and 9 PCA-derived components respectively. As the regularization strength decreased, the accuracies for all but the third experiment generally increased. The third experiment’s SVM model performed slightly better with greater regularization strength (Fig \ref{svm}, \ref{svm_confusion}).

@@

A classification model using Linear Discriminant Analysis (LDA) was also tested across the four experiments. The training data was transformed using Principal Component Analysis (PCA) with the number of components varied between 1 and 12 (Fig \ref{lda}). Using the raw color data from the first experiment, a PCA projection of the data from 12 to 11 components resulted in no loss in test accuracy ($100\%$). Concerning the second experiment, the LDA model achieved its highest accuracy of $91\%$ using 12 components. Again, the addition of color rotation to the data in the third and fourth experiments reduced the accuracy of the model to $77\%$ and $72\%$, respectively.

@@

Using a kNN classification model on the unscaled data across the four experiments, classification accuracies $92\%$, $84\%$, $79\%$, and $50\%$ were achieved for $k=1$, $1$, $1$, $6$, respectively (Fig \ref{knn}, \ref{knn_confusion}). After scaling and transforming the data with PCA, classification accuracies $88\%$, $87\%$, $78\%$, and $70\%$ were achieved for $k=5$, $1$, $1$, $1$, using $2$, $3$, $8$, and $10$ PCA components, respectively (Fig \ref{knn_pca}). Scaling and dimensionality reduction reduced accuracy on the raw color data but increased accuracy on the hue-rotated data.

@@

Across the four experiments, the regression ordinary least squares models generally performed best with low ridge regularization, where $\texttt{log}_{10}(\lambda) \leq 0$  (Fig \ref{ridge}). The lowest mean squared errors achieved across the four experiments were $0.033$, $0.078$, $0.78$, and $0.88$, respectively. The lowest mean squared error for the regression model ($0.033$) was achieved in the first experiment using the lowest regularization parameter tested ($\lambda = 0.001$). The introduction of random gamma modification slightly increased the error in the second experiment, but the color rotation added in the third and fourth experiments affected the accuracy of the model much more significantly.

@@

Using a kNN regression model on the unscaled data across the four experiments, mean squared errors of $0.034$, $0.109$, $0.292$, and $0.949$ were achieved for $k=2$, $2$, $2$, $3$, respectively (Fig 8). The lowest error on this model nearly as reliable as the lowest error achieved using ridge regression ($0.033$). After scaling and using PCA on the kNN regression model, the mean squared error for both the third and fourth experiments was greatly lowered (Fig 9). For $k=11$, $12$ these models achieved mean squared errors of $0.197$ and $0.392$, a significant improvement over the other regression models. Concerning the first two experiments, which did not employ color rotation, kNN achieved mean squared errors of $0.046$ and $0.074$ with $k=3$, $3$ neighbors respectively.

@@

We also experimented with projecting the extracted color data to lower-dimensional subspaces using PCA before running selected classification models. When applied to the unmodified extracted color data, we see that the different classes cluster well and are nearly separable (Fig 3). The pH test strips used in this experiment contain four distinct panels that each react differently to separate ranges of pH values. The neat clustering found by PCA is likely due to the fact that in general, solutions with higher acidity result in brighter colors across the four panels on the pH test strip, whereas more basic solutions result in darker colors. This is why there appears to be an ordering in Figure 3, where the vertical “bars” of points for each decreasing pH class are organized from left to right. When using the PCA-projected data with the kNN (regression/classification) models, we were able to achieve lower error on the data when hue rotation was applied. This makes sense because kNN assumes that training points of the same class should be separated by smaller distances than training points between classes. This “distance” assumption tends to fall apart as the dimensionality of the data increases but this was not a concern for us since our raw color dataset already captured 12 dimensions while PCA worked to further reduce the number of dimensions. Therefore, for the hue-rotated data, PCA was able to help separate the data into clusters where kNN could be more effective with lower values of $k$.

>>>>

We collected our own data for this study. Mutlu was able to capture the range of pH values comprehensively through the use of adjustable chemical buffers. This allowed them to train on chemical solutions with integer pH values. They also were able to determine ground-truth labels for their data points by using a digitized pH meter. We were more limited in our resources; we had a fixed set of non-adjustable buffers and we took the label on the bottle as the ground-truth label for regression. In total we hand-held a smartphone to capture 10 samples under standard household lighting of pH strips submersed in the following 9 buffers: 2.22, 4.40, 4.47, 5.31, 6.28, 7.00, 7.70, 8.70, 10.0 (Fig. \ref{raw_photo}). These fixed non-integer pH values may be a source of irreducible error for us when we rounded these values to the nearest integer when performing classification. However, we wanted our model to be more robust to noise and to be applicable in a more reasonable environment, where a researcher would not have to use a custom apparatus remove any external sources of light [1, 2]. Mutlu waited for the strips to dry before taking pictures, but we decided to capture the images while they were still wet since we imagine that pH strips are read while wet in practice.

@@

For our study, as Mutlu did, we extract out the RBG values from the four panels, but we do this as our first step and use these original features for our first experiment. In order to extract the colors, we first crop the original image using the box specified from the horizontal (Fig. \ref{pre-process}a) and vertical (Fig. \ref{pre-process}b) Sobel edge-detection filters, as illustrated by the red lines in the images. We then apply a repeatedly-opened, distance eroded binary threshold (Fig. \ref{pre-process}c) to get the center regions of each color square from the original image (Fig. \ref{pre-process}d).

@@

From there, we extract the 12 features by taking the average value RBG values from each of the four color boxes (Fig. \ref{four-colors}a). We then generalize Mutlu’s experiments by artificially applying randomly generated lighting conditions to our data. This way, we do not correct for lighting before training a model like Mutlu does but rather, introduce lighting conditions to our homogeneous dataset so that our model can learn to ignore lighting conditions on its own, rather than having to rely on manual and lighting condition-specific pre-processing methods. Specifically, in our second experiment, we gamma correct each sample with a uniformly chosen coefficient between 0.5 and 1.5 (Fig. \ref{four-colors}b) because this mimics the nonlinear luminescence introduced in captured images [3]. In our third experiment, we perform hue rotation with a uniformly chosen degree between -90 and 90 degrees (Fig. \ref{four-colors}c). In our fourth experiment, we apply the random lighting conditions from both experiments 2 and 3 simultaneously (Fig. \ref{four-colors}d).

@@

For select experiments, we used PCA in order to project our data into a lower-dimensional subspace in order to simplify our models and hopefully increase accuracy. We hyperparameter tuned across different numbers of components for the applicable models. We can see that the classes cluster well, even with projection onto only two axes (Fig. \ref{pca}).

@@

As in the Mutlu paper, we first used Support Vector Machines (SVM) for classification across the four experiments. Classification accuracies of $98\%$, $96\%$, $88\%$, and $79\%$ were achieved using 7, 8, 9, and 9 PCA-derived components respectively. As the regularization strength decreased, the accuracies for all but the third experiment generally increased. The third experiment’s SVM model performed slightly better with greater regularization strength (Fig. \ref{svm}, \ref{svm_confusion}).

@@

A classification model using Linear Discriminant Analysis (LDA) was also tested across the four experiments. The training data was transformed using Principal Component Analysis (PCA) with the number of components varied between 1 and 12 (Fig. \ref{lda}). Using the raw color data from the first experiment, a PCA projection of the data from 12 to 11 components resulted in no loss in test accuracy ($100\%$). Concerning the second experiment, the LDA model achieved its highest accuracy of $91\%$ using 12 components. Again, the addition of color rotation to the data in the third and fourth experiments reduced the accuracy of the model to $77\%$ and $72\%$, respectively.

@@

Using a kNN classification model on the unscaled data across the four experiments, classification accuracies $92\%$, $84\%$, $79\%$, and $50\%$ were achieved for $k=1$, $1$, $1$, $6$, respectively (Fig. \ref{knn}, \ref{knn_confusion}). After scaling and transforming the data with PCA, classification accuracies $88\%$, $87\%$, $78\%$, and $70\%$ were achieved for $k=5$, $1$, $1$, $1$, using $2$, $3$, $8$, and $10$ PCA components, respectively (Fig. \ref{knn_pca}). Scaling and dimensionality reduction reduced accuracy on the raw color data but increased accuracy on the hue-rotated data.

@@

Across the four experiments, the regression ordinary least squares models generally performed best with low ridge regularization, where $\texttt{log}_{10}(\lambda) \leq 0$  (Fig. \ref{ridge}). The lowest mean squared errors achieved across the four experiments were $0.033$, $0.078$, $0.78$, and $0.88$, respectively. The lowest mean squared error for the regression model ($0.033$) was achieved in the first experiment using the lowest regularization parameter tested ($\lambda = 0.001$). The introduction of random gamma modification slightly increased the error in the second experiment, but the color rotation added in the third and fourth experiments affected the accuracy of the model much more significantly.

@@

Using a kNN regression model on the unscaled data across the four experiments, mean squared errors of $0.034$, $0.109$, $0.292$, and $0.949$ were achieved for $k=2$, $2$, $2$, $3$, respectively (Fig. 8). The lowest error on this model nearly as reliable as the lowest error achieved using ridge regression ($0.033$). After scaling and using PCA on the kNN regression model, the mean squared error for both the third and fourth experiments was greatly lowered (Fig. 9). For $k=11$, $12$ these models achieved mean squared errors of $0.197$ and $0.392$, a significant improvement over the other regression models. Concerning the first two experiments, which did not employ color rotation, kNN achieved mean squared errors of $0.046$ and $0.074$ with $k=3$, $3$ neighbors respectively.

@@

We also experimented with projecting the extracted color data to lower-dimensional subspaces using PCA before running selected classification models. When applied to the unmodified extracted color data, we see that the different classes cluster well and are nearly separable (Fig. 3). The pH test strips used in this experiment contain four distinct panels that each react differently to separate ranges of pH values. The neat clustering found by PCA is likely due to the fact that in general, solutions with higher acidity result in brighter colors across the four panels on the pH test strip, whereas more basic solutions result in darker colors. This is why there appears to be an ordering in Figure 3, where the vertical “bars” of points for each decreasing pH class are organized from left to right. When using the PCA-projected data with the kNN (regression/classification) models, we were able to achieve lower error on the data when hue rotation was applied. This makes sense because kNN assumes that training points of the same class should be separated by smaller distances than training points between classes. This “distance” assumption tends to fall apart as the dimensionality of the data increases but this was not a concern for us since our raw color dataset already captured 12 dimensions while PCA worked to further reduce the number of dimensions. Therefore, for the hue-rotated data, PCA was able to help separate the data into clusters where kNN could be more effective with lower values of $k$.
